Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber

### 2018-09-04 manuscript/manuscript.tex

When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our state space, but short enough to resolve the dynamics in which we are interested.
Plotting the implied timescales (ITS) as a function of $\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via,
\begin{equation}
\label{eq:its}
t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}.
\end{equation}

@@

\noindent{}where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model.

@@

\noindent{}where the eigenvalues are indexed in decreasing order. The highest eigenvalue, $\lambda_1$, is unique and is equal to $1$, and its corresponding left eigenvector $\phi_1$ corresponds to the stationary distribution of the system.
The right eigenvector $\psi_1$ is a vector consisting of $1$'s.
The subsequent eigenvalues $\lambda_{i>1}$ are real with absolute values less than $1$ and correspond to dynamical processes within the system.
The right eigenvectors $\psi_i$ each represent a dynamical process (for $i>1$), and the coefficients of the eigenvectors represent the flux into and out of the MSM states that characterizes that process.
The corresponding left eigenvectors $\phi_i$ contain the same information weighted by the stationary distribution.

@@

The tutorial software is currently supported for Python versions $3.5$ and $3.6$ on the operating systems Linux, OSX, and Windows.

@@

For the remainder of this manuscript we will walk through the first notebook (00). In notebook 00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide (Fig.~\ref{fig:io-to-tica}a), consisting of $25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of $0.1$~ns. We present the results obtained in the notebook, thereby providing an example of how results generated using PyEMMA can be integrated into research publications.
The figures that will be displayed in the following are created in the showcase notebook (00) and can be easily reproduced.

@@

\begin{figure}
\includegraphics{figure_2}
\caption{Example analysis of the conformational dynamics of a pentapeptide backbone: (a)~The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{vmd}.
(b)~The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
(c)~The sample density projected onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$ ns shows multiple density maxima and
(d)~the time series of the first two ICs show rare transition events.}
\label{fig:io-to-tica}
\end{figure}

\begin{figure}
\includegraphics{figure_3}
\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
(a)~The convergence behavior of the first four implied timescales indicates that a lag time of $\tau=0.5$ ns is suitable for MSM estimation.
(b) A Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns under the assumption of five metastable states accurately predicts the kinetic behavior on longer timescales.
The solid lines in (a) refer to the maximum likelihood result while the dashed lines show the ensemble mean computed with a Bayesian sampling scheme.
The black line indicates the timescale cutoff and, thus, all ITS withion the grey-shaded area are not necessarily resolved.
In both panels, the (non-grey) shaded areas indicate $95\%$ confidence intervals computed with a Bayesian sampling procedure.}
\label{fig:its-and-ck}
\end{figure}

@@

Here, we utilize the VAMP-2 score, which maximizes the kinetic variance contained in the features~\cite{kinetic-maps}.
We should always evaluate the score in a cross-validated manner to ensure that we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
To choose among three different molecular features relevant to protein structure, we compute the (cross-validated) VAMP-2 score at a lag time of $0.5$~ns and find that backbone torsions contain more kinetic variance than the backbone's heavy atom positions or the distances between them (Fig.~\ref{fig:io-to-tica}b).

@@

Subsequently, we perform TICA~\cite{tica,kinetic-maps} in order to reduce the dimension from the feature space, which typically contains many degrees of freedom, to a lower dimensional space that can be discretized with higher resolution and better statistical efficiency. TICA is a special case of the variational principle~\cite{noe-vac,nueske-vamk} and is designed to find a projection preserving the long-timescale dynamics in the dataset. Here, performing TICA on the backbone torsions at lag time $0.5$ ns yields a four dimensional subspace using a $95\%$ kinetic variance cutoff (note that we perform a $\cos/\sin$-transformation of the torsions before TICA in order to preserve their periodicity).
The sample density projected onto the first two independent components (ICs) exhibits several maxima (Fig.~\ref{fig:io-to-tica}c).
Discrete jumps between the maxima can be observed by visualizing the transformation of the first trajectory into these ICs (Fig.~\ref{fig:io-to-tica}d).
We thus assume that our TICA-transformed backbone torsion features describe one or more metastable processes.

@@

\subsection{MSM estimation and validation}
% When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our reduced space, but short enough to resolve the dynamics in which we are interested.
% Plotting the implied timescales (ITS) as a function of $\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}. The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via
% \begin{equation}
% \label{eq:its}
% t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}.
% \end{equation}
A necessary condition for Markovian dynamics in our reduced space is that the ITS are approximately constant as a function of $\tau$; accordingly, we chose the smallest possible $\tau$ which fulfills this condition within the model uncertainty. The uncertainty bounds are computed using a Bayesian scheme~\cite{ben-rev-msm,noe-tmat-sampling} with $100$ samples.
In our example, we find that the four slowest ITS converge quickly and are constant within a $95\%$ confidence interval for lag times above $0.5$~ns (Fig.~\ref{fig:its-and-ck}a). Using this lag time we can now estimate a (Bayesian) MSM with $\tau=0.5$~ns.

To test the validity of our MSM we perform a Chapman-Kolmogorov (CK) test.
% The CK test compares the right and the left side of the Chapman-Kolmogorov equation
% \begin{equation}
% \label{eq:ck}
% T(k \tau) = T^k(\tau)
% \end{equation}
% where $T$ is the MSM transition matrix. The left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k^\textrm{th}$ power.
Visualizing the full transition probability matrix $T$ is difficult; we therefore coarse-grain $T$ into a smaller number of metastable states before performing the test.
An appropriate number of metastable states can be chosen by identifying a relatively large gap in the ITS plot.
For this analysis, we chose five metastable states.
The CK test (Fig.~\ref{fig:its-and-ck}b) shows a good agreement between reestimation at higher lagtimes (black/solid lines) and higher powers of the original transition matrix (blue/dashed lines).
Thus, it confirms that five metastable states is an appropriate choice and shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the long-timescale behavior of our system within error (blue/shaded area).

@@

In order to interpret the slowest relaxation timescales, we refer to the (right) eigenvectors of the MSM as they contain information about what configurational changes are happening and their timescales.
The first right eigenvector corresponds to the stationary process and its eigenvalue is the Perron eigenvalue $1$.
The second right eigenvector, however, corresponds to the slowest process 
(the eigenvector components are real because of the detailed balance constraint enforced during MSM estimation).
The minimal and maximal components of the second right eigenvector indicate the microstates between which the process shifts probability density.
The relaxation timescale of this exchange process is exactly the corresponding implied timescale, which can be computed from its corresponding eigenvalue using~\eqref{eq:its}.
In the projection onto the first two TICA components, we identify the slowest MSM process as a probability shift between macrostate $\mathcal{S}_1$ and the rest of the system, with macrostates $\mathcal{S}_4$ and $\mathcal{S}_5$ in particular (Fig.~\ref{fig:msm-analysis}c).

@@

\begin{figure}
\includegraphics{figure_6}
\caption{Example analysis of the conformational dynamics of a pentapeptide backbone: (a)~the Trp-1 SASA autocorrelation function yields a weak signal which, however, (b)~can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$. The solid/orange lines denote the maximum likelihood MSM result; the dashed/blue lines and the the shaded areas indicate sample means and $95\%$ confidence intervals computed with a Bayesian sampling procedure.}
\label{fig:msm-exp-obs}
\end{figure}

@@

As an example, here we look at the fluorescence correlation of Trp-1, since this terminal tryptophan is a realistic experimental observable for our pentapeptide system.
In order to compute the fluorescence correlation functions we require a microscopic, instantaneous value of the tryptophan fluorescence for each of the original $75$ MSM microstates.
To approximate the fluorescence signal in our pentapeptide system, we use the mdtraj library~\cite{mdtraj} to compute the solvent accessible surface area (SASA) of Trp-1.
Now that we have an approximation of the fluorescence in each of our MSM states, we can use PyEMMA to compute the fluorescence autocorrelation function (ACF) from our MSM (\ref{fig:msm-exp-obs}a).
Note how the computed ACF has a very small response (i.e., signal amplitude).

Using PyEMMA, we can simulate the relaxation of an observable if we had prepared our molecular system in a nonequilibrium initial condition.
The experimental counterpart of such a prediction could be a temperature or pressure jump experiment or a stopped flow assay.
To illustrate such an experiment, we initialize our molecular ensemble as the metastable distribution of $\mathcal{S}_1$ and follow the predicted fluorescence signal as it relaxes to equilibrium (\ref{fig:msm-exp-obs}b).
We see that the predicted relaxation signal has a much larger amplitude for the nonequilibrium initialization, making it more likely to be experimentally measurable.

>>>>

When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our state space, but short enough to resolve the dynamics in which we are interested.
Plotting the implied timescales (ITS) as a function of~$\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via,
\begin{equation}
\label{eq:its}
t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}.
\end{equation}

@@

\noindent{}where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than~1, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model.

@@

\noindent{}where the eigenvalues are indexed in decreasing order. The highest eigenvalue, $\lambda_1$, is unique and is equal to $1$, and its corresponding left eigenvector $\phi_1$ corresponds to the stationary distribution of the system.
The right eigenvector $\psi_1$ is a vector consisting of~$1$'s.
The subsequent eigenvalues $\lambda_{i>1}$ are real with absolute values less than~$1$ and correspond to dynamical processes within the system.
The right eigenvectors $\psi_i$ each represent a dynamical process (for $i>1$), and the coefficients of the eigenvectors represent the flux into and out of the MSM states that characterizes that process.
The corresponding left eigenvectors $\phi_i$ contain the same information weighted by the stationary distribution.

@@

The tutorial software is currently supported for Python versions~$3.5$ and~$3.6$ on the operating systems Linux, OSX, and Windows.

@@

For the remainder of this manuscript we will walk through the first notebook (00).
In notebook~00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide (Fig.~\ref{fig:io-to-tica}a), consisting of~$25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of~$0.1$~ns.
We present the results obtained in this notebook, thereby providing an example of how results generated using PyEMMA can be integrated into research publications.
The figures that will be displayed in the following are created in the showcase notebook (00) and can be easily reproduced.

@@

\begin{figure}
\includegraphics{figure_2}
\caption{Example analysis of the conformational dynamics of a pentapeptide backbone: (a)~The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{vmd}.
(b)~The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
(c)~The sample density projected onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$~ns shows multiple density maxima and
(d)~the time series of the first two ICs of the first trajectory.}
\label{fig:io-to-tica}
\end{figure}

\begin{figure}
\includegraphics{figure_3}
\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
(a)~The convergence behavior of the implied timescales associated with the four slowest processes.
(b)~Chapman-Kolmogorov test computed using an MSM estimated with lag time $\tau=0.5$~ns assuming~5 meta-stable states..
The solid lines in (a) refer to the maximum likelihood result while the dashed lines show the ensemble mean computed with a Bayesian sampling procedure~\cite{ben-rev-msm}.
The black line indicates where implied timescales are equal to the lag time, whereas the grey area indicates all implied timescales faster than the lag time.
In both panels, the (non-grey) shaded areas indicate~$95\%$ confidence intervals computed with the aforementioned Bayesian sampling procedure.}
\label{fig:its-and-ck}
\end{figure}

@@

Here, we utilize the VAMP-2 score, which maximizes the kinetic variance contained in the features~\cite{kinetic-maps}.
We should always evaluate the score in a cross-validated manner to ensure that we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
To choose among three different molecular features relevant to protein structure, we compute the (cross-validated) VAMP-2 score at a lag time of~$0.5$~ns.
We find that backbone torsions contain more kinetic variance than the backbone heavy atom positions or the distances between them (Fig.~\ref{fig:io-to-tica}b).
This suggests that backbone torsions are the best of the options evaluated for MSM construction.

@@

Subsequently, we perform TICA~\cite{tica,kinetic-maps} in order to reduce the dimension from the feature space, which typically contains many degrees of freedom, to a lower dimensional space that can be discretized with higher resolution and better statistical efficiency.
TICA is a special case of the variational principle~\cite{noe-vac,nueske-vamk} and is designed to find a projection preserving the long-timescale dynamics in the dataset.
Here, performing TICA on the backbone torsions at lag time~$0.5$~ns yields a four dimensional subspace using a~$95\%$ kinetic variance cutoff (note that we perform a $\cos/\sin$-transformation of the torsions before TICA in order to preserve their periodicity).
The sample density projected onto the first two independent components (ICs) exhibits several maxima (Fig.~\ref{fig:io-to-tica}c).
Discrete jumps between the maxima can be observed by visualizing the transformation of the first trajectory into these ICs (Fig.~\ref{fig:io-to-tica}d).
We thus assume that our TICA-transformed backbone torsion features describe one or more metastable processes.

@@

\subsection{MSM estimation and validation}
% When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our reduced space, but short enough to resolve the dynamics in which we are interested.
% Plotting the implied timescales (ITS) as a function of~$\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}. The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via
% \begin{equation}
% \label{eq:its}
% t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}.
% \end{equation}
A necessary condition for Markovian dynamics in our reduced space is that the ITS are approximately constant as a function of $\tau$; accordingly, we chose the smallest possible $\tau$ which fulfills this condition within the model uncertainty. The uncertainty bounds are computed using a Bayesian scheme~\cite{ben-rev-msm,noe-tmat-sampling} with~$100$ samples.
In our example, we find that the four slowest ITS converge quickly and are constant within a $95\%$ confidence interval for lag times above~$0.5$~ns (Fig.~\ref{fig:its-and-ck}a). Using this lag time we can now estimate a (Bayesian) MSM with $\tau=0.5$~ns.

To test the validity of our MSM we perform a Chapman-Kolmogorov (CK) test.
% The CK test compares the right and the left side of the Chapman-Kolmogorov equation
% \begin{equation}
% \label{eq:ck}
% T(k \tau) = T^k(\tau)
% \end{equation}
% where $T$ is the MSM transition matrix. The left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k^\textrm{th}$ power.
Visualizing the full transition probability matrix $T$ is difficult; we therefore coarse-grain $T$ into a smaller number of metastable states before performing the test.
An appropriate number of metastable states can be chosen by identifying a relatively large gap in the ITS plot.
For this analysis, we chose five metastable states.
The CK test (Fig.~\ref{fig:its-and-ck}b) shows that predictions from our MSM (blue-dashed lines) agrees well with MSMs estimated with longer lag times (black-solid lines)
Thus, the CK test confirms that five metastable states is an appropriate choice and shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the long-timescale behavior of our system within error (blue/shaded area).

@@

In order to interpret the slowest relaxation timescales, we refer to the (right) eigenvectors of the MSM as they contain information about what configurational changes are happening and their timescales.
The first right eigenvector corresponds to the stationary process and its eigenvalue is the Perron eigenvalue~$1$.
The second right eigenvector, however, corresponds to the slowest process 
(the eigenvector components are real because of the detailed balance constraint enforced during MSM estimation).
The minimal and maximal components of the second right eigenvector indicate the microstates between which the process shifts probability density.
The relaxation timescale of this exchange process is exactly the corresponding implied timescale, which can be computed from its corresponding eigenvalue using~\eqref{eq:its}.
In the projection onto the first two TICA components, we identify the slowest MSM process as a probability shift between macrostate $\mathcal{S}_1$ and the rest of the system, with macrostates $\mathcal{S}_4$ and $\mathcal{S}_5$ in particular (Fig.~\ref{fig:msm-analysis}c).

@@

\begin{figure}
\includegraphics{figure_6}
\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
(a)~the Trp-1 SASA autocorrelation function yields a weak signal which, however,
(b)~can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$.
The solid/orange lines denote the maximum likelihood MSM result;
the dashed/blue lines and the the shaded areas indicate sample means and~$95\%$ confidence intervals computed with a Bayesian sampling procedure~\cite{ben-rev-msm}.}
\label{fig:msm-exp-obs}
\end{figure}

@@

As an example, here we look at the fluorescence correlation of Trp-1, since this terminal tryptophan is a realistic experimental observable for our pentapeptide system.
In order to compute the fluorescence correlation functions we require a microscopic, instantaneous value of the tryptophan fluorescence for each of the original~$75$ MSM microstates.
To approximate the fluorescence signal in our pentapeptide system, we use the mdtraj library~\cite{mdtraj} to compute the solvent accessible surface area (SASA) of Trp-1.
Now that we have an approximation of the fluorescence in each of our MSM states, we can use PyEMMA to compute the fluorescence autocorrelation function (ACF) from our MSM (\ref{fig:msm-exp-obs}a).
Note how the computed ACF has a very small response (i.e., signal amplitude).

Using PyEMMA, we can simulate the relaxation of an observable if we had prepared our molecular system in a nonequilibrium initial condition.
The experimental counterpart of such a prediction could be a temperature or pressure jump experiment or a stopped flow assay.
To illustrate such an experiment, we initialize our molecular ensemble as the metastable distribution of~$\mathcal{S}_1$ and follow the predicted fluorescence signal as it relaxes to equilibrium (\ref{fig:msm-exp-obs}b).
We see that the predicted relaxation signal has a much larger amplitude for the nonequilibrium initialization, making it more likely to be experimentally measurable.
